---
title: "gdx visulizer"
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---

## 0. Settings

```{r, warning=FALSE, message=FALSE}
rm( list = ls() )
list.files()
graphics.off() # chiude tutti i device grafici
cat("\014") #pulisci console

library(tidyverse)
library(gdxtools)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(ggsci) # scientific colors
library(viridis) #looks good and includes many options
library(RColorBrewer) # wide choice
library(countrycode) # Country name tools

```

## 1. gdx

```{r}
mygdx <- gdx("results_ssp2_bau.gdx")
var_names = mygdx$variables$name
par_names = mygdx$parameters$name
sets_names = mygdx$sets$name
```

### simple plot

```{r}
temp_usa <- subset(mygdx["TEMP"], 
                   n == "usa" & m == "atm", 
                   select = c("t", "value"))
temp_usa
plot(temp_usa,lwd=2,type = "l")
```

### To get a vector of the global regions from the set n:

```{r}
mygdx["n"][[1]]
mygdx["n"]
```

### To get a data.frame of the parameter wemi:

```{r}
mygdx["wemi"]
mygdx["wemi"] %>% as_tibble() # removes row names and each colum is a list?
```

### The definition of the symbol can be available with its attribute "gams".

```{r}
print(attr(mygdx["wemi"],"gams"))
```

### to get a data.frame of the variable Q_EMI (its level by default)

```{r}
mygdx["Q_EMI"] %>% as_tibble()
# to load upper bound
mygdx["Q_EMI",field = "up"] %>% as_tibble()
```

### To load many gdx at once in one data.frame, use the function batch_extract

```{r}
myfiles = file.path('C:/Users/modin/Desktop/Ettore/UNIVERSITA/ECC_GAMS/R STUDIO',c("results_ssp2_bau.gdx","results_ssp2_ctax50.gdx"))
qemi = batch_extract("Q_EMI",files = myfiles)[[1]]
qemi %>% as_tibble()
#mygdx["TEMP"]$value

```

### load CSV or other datas

```{r}
dice_csv <- read_csv('Dice2016R-091916ap.csv', skip = 6) # returns already a tibble
# rename Year to Variable
dice_csv <- dice_csv%>%rename(Variable=Year)
dice_csv
# let's make a long table we will need it for plotting
mdat_2c = dice_csv %>% pivot_longer(!Variable, names_to = "Year")
# Year should be numerical
mdat_2c <- mdat_2c %>% mutate(Year = as.numeric(Year)) 
mdat_2c
```

## 2. plot graphs

```{r}
ggplot(data = mdat_2c %>% filter(Variable == 'Consumption'),
        mapping = aes(x = Year,y = value)) +
geom_line() +
labs(title = "Consumption - DICE - 2C Scenario", x = "", y = "Consumption [trillion USD]") +
theme_bw() +
scale_y_continuous(expand = c(0,0), limits = c(0,NA)) # Includes 0 in the y-axis, NA takes the max in the dataset
# + scale_x_continuous(expand = c(0,0), limits = c(2020,2320)) # Zoom-in in the x-axis

```

```{r}
vars = c('Consumption','Atmospheric Temperature',
'Industrial Emissions GTCO2 per year',
'damages')
p<-ggplot(mdat_2c %>% filter(Variable %in% vars),
        aes(x = Year, y = value)) +
geom_line() +
labs(x = "", y = "") +
facet_wrap(~Variable, ncol = 4, scales='free') + # - free scales
theme_bw()
p

```

### save graphs

```{r}
#ggsave(filename = "Fig1.pdf", plot = p, width = 21, height = 29.7, units = "cm") # A4 format
#ggsave(filename = "Fig1.png", plot = p, width = 8, height = 5) # smaller png file
```

## 3. 2D graphs

```{r}
# read several GDX (several scenarios with different carbon taxes)
myfiles <- file.path("C:/Users/modin/Desktop/Ettore/UNIVERSITA/ECC_GAMS/R STUDIO", c("results_ssp2_bau.gdx","results_ssp2_ctax50.gdx","results_ssp2_ctax100.gdx"))
res <- batch_extract("Q_EMI", files = myfiles) # load GHG emissions
# have a look at the structure
str(res) # a list of a data frame

# this grabs the data.frame inside the list GHG$Q_EMI
qemi = res$Q_EMI %>% as_tibble()
# The element in a list can also be accessed with res[[1]] or res[['Q_EMI']]
qemi

qemi_brazil = qemi[which(qemi$n=="brazil"),]
qemi_brazil

```

### transformations

```{r}
qemi <- qemi %>%
  mutate(year = as.numeric(t) * 5 + 2000) %>% # Transform time period into year
  mutate(scen = basename(gdx)) %>% # Transform the filename into a scenario name
  mutate(scen = str_replace(scen, ".gdx", "")) %>%
  mutate(scen = str_replace(scen, "results_ssp2_", ""))
  # basename returns the filename without the path
  # str_replace replace a given string pattern

```

#### Update the scenario as factors (ordered set)

```{r}
qemi <- qemi %>%
  mutate(scen = factor(scen, levels = unique(qemi$scen)))

qemi
```

### We have 4 id variables, let's reduce it to 3 by aggregation regions into world

```{r}
# group_by scenarious, emission and year and summurise other all countrys using sum
wemi <- qemi %>%
group_by(scen,e,year) %>%
summarise(value = sum(value))

wemi

# Let's change unit from GtC to GtCO2
wemi <- wemi %>%
mutate(value = value * 44 / 12)



ggplot(wemi %>% filter(e == 'co2' & year <= 2100),
      aes(x = year, y = value, color = scen)) +
  
geom_line() +
  
geom_point() + # NEW GEOMETRY, SAME DATA AND MAPPING
  
#geom_point(data = wemi %>% filter(e = 'co2' & year = 2100 & scen = "ctax100")) + #SUBSET BUT SAME MAP
  
geom_hline(yintercept = 0, color = 'Black', size = 0.8) +
  
labs(x = "", y = "CO2 [GtC]", title = "CO2 emissions") +
  
theme_bw() +
  
scale_x_continuous(limits=c(2005, 2100),
                    breaks=seq(2005, 2100,by = 10), # Define the axis tiks breaks
                    guide = guide_axis(n.dodge = 2)) + # avoid overlapping
  
scale_y_continuous(limits=c(-5, 50)) +
  
#scale_color_discrete(name = "Scenario")

scale_color_viridis(discrete = TRUE, option = "D") # use "?" to check the options
```

## 4. Scatter

### plots-load more data

```{r}
pop <- batch_extract("l",files = myfiles)[[1]] %>% as_tibble() # Load population
# let's transform time periods into Years
pop <- pop %>%
  mutate(year = as.numeric(t) * 5 + 2000) %>% # Transform time period into year
  mutate(scen = basename(gdx)) %>% # Transform the filename into a scenario name
  mutate(scen = str_replace(scen, ".gdx", "")) %>%
  mutate(scen = str_replace(scen, "results_ssp2_", ""))

```

### Merging two data.table objects

```{r}
# let's keep 2 emission parameters
dat = full_join(qemi %>% filter(e %in% c('kghg','ccs')) %>% select(-gdx,-t),
pop %>% select(-gdx,-t),
by = c("scen","year","n"),
suffix = c(".ghg",".pop"))
# have a look, it's magical
dat

```

### adding dimensions

```{r}
ggplot(dat %>% filter(year < 2100 & e == 'kghg'),
aes(x = value.pop, y = value.ghg, color = n)) +
geom_point()

```

### advanced options

```{r}
ggplot(dat %>% filter(year < 2100 & e == 'kghg'),
aes(x = value.pop,y = value.ghg, color = n, 
          sahpe = scen,    # add shape
          alpha = year,    # add transparency
          size = e)) +     # add size with 'e'
geom_hline(yintercept = 0) + # add a horizontal line at 0
geom_vline(xintercept = 1000, color = 'red') + # Separate low-high population regions
geom_point() +
geom_point(data = dat %>% filter(year == 2050 & value.pop > 1000 & e == 'kghg'),
            color='green')+ # Emphasize a few points

geom_label_repel(aes(label = n),
                dat %>% filter(year == 2050 & value.pop > 1000 & e == 'kghg'),
                color = 'black', alpha = 1) + # Add labels
scale_x_continuous(name = "Population [millions]",expand = c(0.1, 0.1)) +
scale_y_continuous(name = "CO2 [GtCe]",expand = c(0.1, 0.1))+
scale_color_viridis(discrete = TRUE, option = "B", name = "Region") +
facet_grid(e~scen, scales='free') + # Add facets
guides(col = guide_legend(ncol = 9)) + # Legend in 9 columns
#theme_bw() +
theme(legend.position = "bottom") # Legend in the bottom

```

## 5. Box plots Plotting distribution

```{r}
# Update the scenario as factors (ordered set)
dat <- dat %>% mutate(scen = factor(scen, levels = c('bau','ctax50','ctax100')))

ggplot(dat %>% filter(e == 'kghg'), aes(x = scen, y = value.ghg)) +
geom_boxplot(outlier.shape = NA) + theme_bw() +
geom_hline(yintercept = 0, color = "red") +
geom_jitter(aes(color = n), shape = 16,
          position = position_jitter(0.2), alpha = 0.6) + # Add observations
stat_summary(fun = mean, geom = "point", shape = 16, size = 4) + # Add mean distribution
labs(x = "Scenario", y = "GHG [GtCe]") +
scale_color_viridis(name="Regions",discrete = TRUE, option = "D")+
guides(col = guide_legend(ncol = 3))# color per region
```

## 6. Maps - Showing regional information

```{r}
# import the WITCH model regional definition
wreg <- read_csv('mapwitch17.csv')
# transforms ISO codes into country names
wreg <- wreg %>% mutate(region = countrycode(ISO, origin = 'iso3c', destination = 'country.name'))
# Correcting some mismatches
#wreg[ISO ='GBR']$region = 'UK'
#wreg[ISO ='USA']$region = 'USA'
wreg

world <- ne_countries(scale = "small", returnclass = "sf")
world <- subset(world,!adm0_a3 %in% c("ATA","FJI"))
# merge the WITCH regional definition with the world map
world <- merge(world,wreg, by.x = "adm0_a3", by.y = "ISO")
# merge with the emission data.frame
world0 <- merge(world,qemi %>% filter(year == 2100 & e == "co2"), by = "n", allow.cartesian=TRUE)
# Use a better projection 'equal projection'
target_crs <- '+proj=eqearth +wktext'
world1 <- st_transform(world0, crs = target_crs)
```

### showing regional information

```{r}
#ggplot(data = world1 %>% filter(scen == "ctax50")) +
ggplot(data = world1) +
      geom_sf(aes(fill = value)) +
      coord_sf(datum = target_crs, expand = FALSE, clip = "off") +
      scale_fill_distiller(name = "CO2 Emissions\nin 2100", palette ="RdYlGn",direction=-1) +
theme_void() +
theme(legend.position = "bottom",
      strip.text.x = element_text(size = 12, face="bold"))+
facet_grid(~scen)
```

### trasforming variables for deeper analysis

```{r}
# Put the scenarios values in columns
dqemi <- qemi %>%
select(-t,-gdx) %>%
filter(e == "co2" & year == 2100) %>%
pivot_wider(names_from = "scen", values_from = "value")
# Let's create a new variable/column
dqemi <- dqemi %>% mutate(dif_wrtBAU = bau - ctax100)
# merge with the emission data.frame
world2 <- merge(world,dqemi, by = "n", allow.cartesian=TRUE)
# Use a better projection 'equal projection'
target_crs <- '+proj=eqearth +wktext'
world3 <- st_transform(world2, crs = target_crs)

ggplot(data = world3) +
geom_sf(aes(fill = dif_wrtBAU * 44 / 22)) +
coord_sf(datum = target_crs, expand = FALSE, clip = "off") +
scale_fill_distiller(name = "CO2 emission\nreduction\n[GtCO2]",
                     palette ="RdYlGn", direction = -1) +
theme_void()

```
